White Dwarf Mergers on Adaptive Meshes\\ II. Collisions

### 2016-02-14 papers/wdmerger_II/paper.tex
Update section on timestep limiting in paper II

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$t_e$:
\begin{equation}
  t_e \equiv \frac{e}{|\dot{e}|},
\end{equation}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, we define a burning-limited timestep $\Delta t_{b}$:
\begin{equation}
  \Delta t_{b} = f_{\Delta t}\, t_e.
\end{equation}
We estimate $\dot{e}$ using the energy change in the last burning
update, divided by the length of the burning timestep. The factor $f_{\Delta t}$ then
determines by what fraction we would like to allow the internal energy to change
in the current timestep (assuming that $\dot{e}$ does not change quickly from
timestep to timestep, which is not always the case). By making $f_{\Delta t}$ smaller,
we can control the magnitude of the decoupling between the reactions and the hydro.
We choose by default $f_{\Delta t} = 0.1$, a choice which is on par with others in 
the literature. The sensitivity of results to this choice will be discussed in 
\autoref{sec:2D:timestep}. The factor $f_{\Delta t}$ can be set at runtime in \castro.

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of the
burning timestep across all zones.

\citet{kushnir:2013} point out this is not the only way for the burning to
become numerically unstable; another failure mode is when the energy injection timescale
$t_e$ is shorter than the sound-crossing time $t_s$ in a zone.
The goal is to ensure that the following condition holds:
\begin{equation}
  t_s \leq f_{\Delta x} t_e \label{eq:burning_limiter_2}
\end{equation}
The sound crossing time, $t_s$, is given by $\Delta x / c_s$, 
where $c_s$ is the sound speed and $\Delta x$ is the (minimum) 
zone width. The parameter $f_{\Delta x}$ then determines the minimum
ratio of the nuclear energy generation timescale to the 
sound-crossing time. \citet{kushnir:2013} choose $f_{\Delta x} = 0.1$ 
for their simulations, and we do too (this parameter can be set 
at runtime in \castro).

@@

While this approach may add several AMR levels 
to a simulation, we agree with \citet{kushnir:2013} that 
solving this numerical instability is crucial to avoiding
early, unphysical detonations, and so it is unavoidable
if a correct evaluation of the burning phase is desired.
A simulation that does not solve this problem in this way
(or some analogous manner) will not obtain the correct amount 
of burning, and will not converge properly with resolution. 
Fortunately, the regions where this criterion is not satisfied 
are fairly localized on the grid, so the amount of additional 
computational work is not exceedingly large.

>>>>

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{e}\, \tau_e \\
  \Delta t_{bX_k} &= f_{X}\, \tau_{X_k}.
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{e}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{X}$ 
determines the maximum change in the mass fraction of any species. By making 
$f_{e}$ and $f_{X}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. We choose by default $f_{X} = f_{e} = 0.1$, 
a choice which is slightly smaller than others in the literature. (Note that previous 
collision papers we are aware of only limited based on changes in internal energy, 
not on cahnges in species.)  The sensitivity of results to this choice will be discussed in 
\autoref{sec:2D:timestep}. The factors $f_{e}$ and $f_{X}$ can be set at runtime in \castro.

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of all the
burning timesteps across all zones. To do this, we need a method for determining 
$\dot{e}$ and $\dot{X_k}$. A typical choice in the literature has been to set, for example,
\begin{equation}
  \dot{e} = \frac{e^{n} - e^{n-1}}{\Delta t^{n-1}}, \label{eq:burning_mode_4}
\end{equation}
where is $e^n$ is the internal energy at the start of the current timestep and
$e^{n-1}$ is the internal energy at the start of the previous timestep. 
The obvious analogue is used for constructing the species rate of change.
However, there are alternative methods of constructing this derivative estimate, 
and we have found that these different methods have measurable consequences.
We define four separate methods for calculating the time derivative, with 
the above being method 4. Method 3 is similar to method 4 but replaces the
denominator in Equation \ref{eq:burning_mode_4} with the change in 
internal energy over the last timestep only from the nuclear reactions.
Method 2 is the same as method 3 but we only use the change in internal 
energy from the most recent nuclear burning step, that is, the second-half
of the Strang-split burning from the last timestep (the denominator 
then becomes $\Delta t / 2$). In method 1, the most accurate option and 
the current default in \castro, we evaluate the right-hand-side of the 
burning network given the current state to explicitly obtain the 
instantaneous value of $\dot{e}$ and $\dot{X_k}$.

To understand the consequences of this choice, and more broadly to 
understand the limitations of Strang splitting, we consider the 
basic outline of a single-level advance in an advection-reaction system:
\begin{enumerate}
  \item Evaluate timestep $\Delta t$ for the current advance
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Advance the hydrodynamics by $\Delta t$
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Return to Step 1
\end{enumerate}
Now, consider that during a head-on collision, initial nuclear burning 
will occur at the contact point between the two stars. Because of 
the staggered updates from splitting, the evolution effectively progresses 
as a cycle between burning for $\Delta t$ and getting fresh material 
advected into the contact point by the hydro update for $\Delta t$. 
When the collision begins, $\Delta t$ is controlled by the hydrodynamic 
stability criterion, and may be large enough that it is possible for 
the burning advance in Step 4 to completely burn the freshly advected 
material all the way to NSE. Consequently the evolution is no longer 
a good approximation to smooth burning of the in-falling material but
rather separate discrete burning and hydro steps, and the nature of 
the burning evolution will be quite different. Furthermore, by the 
time we return to Step 1 and estimate the next timestep size, all 
of the burning rates will be small again, and the instantaneous 
timestep limiter of method 1 may actually substantially overestimate 
the needed timestep. The other methods will see that the energy/species  
substantially changed over the last timestep, but will still 
overestimate the needed timestep because the burning was quiescent
for at least some portion of the last advance. In particular, our 
experience has shown that using only the energy change criterion
is particularly susceptible to this phenomenon; silicon-group 
material can build up without changing the internal energy by a 
large fraction, so the timestep limiter is never triggered, and 
then in a single step a substantial amount of iron-group elements 
can be be generated. This can have non-trivial effects on the 
total amount of iron-group elements generated over the course of
the simulation. We have found that the addition of the 
limiter based on changes in species functionally precludes this.

With the timestep limited the way we advocate in this paper, 
the timesteps are generally short enough so that the errors 
due to splitting are small. Other approaches to the coupling 
between reactions and hydrodynamics have been proposed in the 
broader literature, especially iterative methods such as 
deferred corrections that allow each of these operators to 
feel the lagged effects of the other operators. For example,
in the context of low Mach number flows, \cite{nonaka:2012} have
used the method of spectral deferred corrections \cite{SDC} to
couple their advection-diffusion-reaction equation set. In our
context this would involve using the advection as a source term
for the burning network, and symmetrically including a source
term to the hydrodynamics corresponding to the energy release
from the burning. We are presently investigating such a method,
and it may form the basis of further work on this subject.

Finally, we return to a point we hinted at above. The timestep
will only actually satisfy the energy criterion
$\Delta t \leq f_e \tau_e$ and species criterion
$\Delta t \leq f_{X_k} \tau_{X_k}$ when the estimates for
$\dot{e}$ and $\dot{X_k}$ we generate are at least as large
as the actual rate of change of energy and mass fractions
over the timestep. However, this can assumption can fail
quite radically during periods of runaway burning when
the rate of change of these quantities is highly nonlinear.
We do not want to neglect the errors caused by this approximation
because they may build up over an extended period of nonlinear
evolution and perhaps substantially change the final results.
To this end, we have implemented a timestep retry option in
\castro. The retry option works by first saving a copy of the
full state, and then doing the advance. We then explicitly
check whether the timestep criteria were violated during the advance,
using the time-centered energy and mass fractions and a
time-centered differencing of the old and new states to
calculate the evolution timescale. If so, we reject this
advance and restore the original state data. We then perform
a subcycled update, with the maximum timestep chosen so that
the criterion is satisfied in each subcycle. Note that this
occurs entirely on a single level and is separate from the
subcycling that is used by \boxlib\ to advance the different
AMR levels in time; the full AMR algorithm does not ever
know that the retry subcycling took place. This can help
substantially to mollify the problem described above, where
the time derivative estimation methods proposed were not
sufficient to obtain the desired limiting. Thus we strongly
recommend that hydrodynamics codes implement such an approach,
which ensures that the coupling actually does stay at the
level requested. The most significant drawback is that saving
a copy of the old state effectively doubles the memory footprint
of the code, so this option should be disabled when not needed.
However, it may be crucial in determining the evolution of
systems such as white dwarf collisions, where even a short
period of incorrect advances can significantly pollute the
final answer, and the additional memory expense is justified
by obtaining a more accurate solution.

\subsection{Numerically Unstable Burning}

\citet{kushnir:2013} point out that an inappropriate timestep is 
not the only way for the numerical discretization to cause 
severe errors in the burning. Another failure mode is when
the energy injection timescale
$t_e$ is shorter than the sound-crossing time $t_s$ in a zone.
The goal is to ensure that the following condition holds:
\begin{equation}
  t_s \leq f_{s} t_e \label{eq:burning_limiter_2}
\end{equation}
The sound crossing time, $t_s$, is given by $\Delta x / c_s$, 
where $c_s$ is the sound speed and $\Delta x$ is the (minimum) 
zone width. The parameter $f_{s}$ then determines the minimum
ratio of the nuclear energy generation timescale to the 
sound-crossing time. \citet{kushnir:2013} choose $f_{s} = 0.1$ 
for their simulations, and we do too (this parameter can be set 
at runtime in \castro).

@@

While this approach may add several AMR levels 
to a simulation, we agree with \citet{kushnir:2013} that 
solving this numerical instability is crucial to avoiding
early, unphysical detonations, and so it is unavoidable
if a correct evaluation of the burning phase is desired.
A simulation that does not solve this problem in this way
(or some analogous manner) will not obtain the correct amount 
of burning, and will not converge properly with resolution. 
Fortunately, the regions where this criterion is not satisfied 
are fairly localized on the grid, so the amount of additional 
computational work is not insurmountable for a single simulation.
However, this level of accuracy may make a parameter study of
the problem unfeasible, given current computational resources.
